Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.

### 2023-06-01 scripts/manuscript/manuscript.qmd
secondary sampling scheme, revised manuscript for ncc

---
title: Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
abstract: |
  [insert abstract here - nicholas ignore the formatting on the authors for now, i can output to pdf and have it formatted nicely for submission]
keywords: 
  - structural equation modelling
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Remote Sensing of Environment
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
---

@@

# include all libraries required by the R chunks here
library(vegan)
library(sgsR)
library(terra)
library(sf)

@@

# generate package citations
loaded_pkgs = (.packages())
base_pkgs = getOption("defaultPackages")
contrib_pkgs = loaded_pkgs[!(loaded_pkgs %in% base_pkgs)]
pkgs_file = here::here("scripts", "manuscript", "packages.bib")
knitr::write_bib(contrib_pkgs, pkgs_file)
```

@@

Biodiversity is the summation of variation in biological life, across
genes, species, communities, and ecosystems. Currently, biodiversity is
in decline, facing extinction rates above the background extinction rate
[@thomas2004; @urban2015], and homogenization of communities at various
scales [@mcgill2015]. In response, the global biodiversity community is
making efforts to assess and halt the degradation of biodiversity on
earth. The Group for Earth Observation Biodiversity Observation Network
has developed the Essential Biodiversity Variables [EBVs, @pereira2013],
designed as an analog to the Essential Climate Variables framework
[@bojinski2014]. EBVs are designed to be global in scope, relevant to
biodiversity information, feasible to utilize, and complementary to one
another [@skidmore2021]. There are six EBV classes, each of which
correspond to a different facet of biodiversity including species
populations, species traits, community composition, ecosystem structure,
ecosystem function, and genetic composition [@pereira2013].

Satellite remote sensing has proven to be capable of measuring five of
the six EBV classes, the exception being genetic composition, which
requires in-situ observation and sampling [@skidmore2021]. Species
populations - and in turn community composition - can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale, however it is difficult and computationally expensive to extend
these analyses to broader extents [@fassnacht2016; @graves2016] while
species traits such as vegetation phenology have been observed at the
single-tree scale using, for example, PlanetScope imagery and
drone-based measurements [@wu2021]. Coarser spatial measurements, such
as those taken from satellites by the MODIS [@zhang2003], Landsat
imaging systems[@fisher2006], or Sentinel [@helfenstein2022;
@darvishzadeh2019] can monitor these functional processes at broader
extents, corresponding to ecosystem function, but the coarse spatial
resolution removes the ability to relate these traits to individual
trees. Remote sensing data has been shown to be arguably the most
effective at monitoring ecosystem structure and function at regional to
global extents through the use of optical imagery [@cohen2004] and
active sensors such as lidar and radar datasets [@lefsky2002; @lang2021;
@neuenschwander2019].

Notably, while remote sensing can provide information on the
species-level EBVs, this information is typically acquired with an
*ad-hoc* approach requiring high spatial and spectral resolution data
[@skidmore2021]. This data can often only be collected at small extents,
rather than the global or regional scales required for biodiversity
trend assessment [@valdez2023]. The two landscape-level EBVs (ecosystem
structure and function) are well suited to be examined at global or
regional scales using moderate-resolution satellite imagery, such as
those provided by the Landsat or Sentinel series of satellites
[@skidmore2021].

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Many metrics derived from
lidar remote sensing have been used as local indicators of biodiversity,
including simple metrics such as canopy cover and canopy height as well
as derived metrics including vertical profiles, aboveground biomass
[@lefsky1999]. Other second order derived metrics such as canopy
texture, height class distribution, edges, and patch metrics have also
been used to examine habitat and biodiversity at landscape scales
[@bergen2009]. Increased structural complexity has been hypothesized to
create additional niches, leading to increased species diversity
[@bergen2009], which has been frequently demonstrated using avian
species diversity metrics [@macarthur1961]. For example: @herniman2020
used spectral and lidar derived forest structure data to model avian
habitat suitability; @clawges2008 found that lidar derived forest
structural attributes are capable of identifying habitat types
associated with avian species in pine/aspen forests; @goetz2007 used
canopy structural diversity to predict bird species richness, finding
that canopy vertical distribution was the strongest predictor of species
richness. Forest structural metrics have also been used to study
biodiversity in other clades as well [@davies2014; @nelson2005].

With respect to ecosystem function, energy availability in an ecosystem
has shown to be a predictor of species richness and abundances at
various scales [@chase2002; @radeloff2019; @coops2019; @razenkova], and
is measurable using satellite remote sensing via the use of various
vegetation indices [@huete2002; @radeloff2019]. Vegetation indices,
which are indicative of photosynthetic activity, are commonly used as
proxies of gross primary productivity [@huang2019]. These vegetation
indices have also been used to assess patterns in biodiversity at single
time points [@bonn2004], and more recently, through yearly summaries of
productivity [@radeloff2019]. The relationship between energy
availability and biodiversity occurs via various hypotheses, such as the
available energy hypothesis, the environmental stress hypothesis, and
the environmental stability hypothesis. These three hypotheses have in
turn been linked to patterns of surface reflectance in remote sensing
data [@radeloff2019]. The Dynamic Habitat Indices (DHIs) have been shown
to be well suited to assess these hypotheses. The cumulative DHI
calculates the total amount of energy available in a given pixel over
the course of a year. Cumulative DHI is strongly linked to the available
energy hypothesis, which suggests that with greater available energy
species richness will increase [@wright1983]. The minimum DHI, which
calculates the lowest productivity over the course of a year can be
matched to the environmental stress hypothesis, which proposes that
higher levels of minimum available energy will lead to higher species
richness [@currie2004]. Finally, the variation DHI, which calculates the
coefficient of variance in a vegetation index through the course of a
year, corresponds to the environmental stability hypothesis which states
that lower energy variation throughout a year will lead to increased
species richness [@williams2008].

Within forested ecosystems, there are therefore many possible remotely
sensed derived ecosystem structure and function metrics that can be
assessed. Structural variables range in complexity from simple (canopy
cover; canopy height), to more complex (elevation coefficient of
variation; elevation standard deviation) to modelled (aboveground
biomass; basal area), all of which can be assessed using lidar data
[@coops2021]. Data on ecosystem functionality ranges from phenology
metrics (date of leaf on/leaf off; length of growing season) to
functional traits (chlorophyll content; leaf area index) to productivity
estimates (GPP; total available energy) [@pettorelli2018]. Advances in
satellite remote sensing processing have allowed 3D forest structure
data to be imputed across wide spatial scales [@matasci2018; @coops2021]
using data fusion approaches involving collected lidar data and
optical/radar data. Other advances in image compositing have allowed
yearly summaries of vegetation productivity to be calculated at regional
to global scales, summarizing the yearly total, minimum, and variation
in energy availability at similar spatial resolutions [@radeloff2019].

Linkages between forest structure and function have been examined within
a remote sensine context for over 20 years [@huete2002, @knyazikhin1998;
@myneni1994]. While there is significant theoretical and empirical
evidence for their relationship at single time points (within a single
image) [@myneni1994], various relationship directions and shapes have
been found between forest structure and function metrics [@ali2019].
These relationships, their shapes, and their strengths have been
attributed to multiple possible hypotheses and have been shown to vary
based on environmental conditions [@ali2019]. The relationship in
particular between forest structural diversity metrics - which are now
more accurately and comprehensively derived from lidar data - and
temporal variation in functional metrics, specifically the metrics of
ecosystem productivity via the DHI framework, have yet to be fully
examined.

Understanding the complementarity between potential EBVs is an integral
component of their creation [@skidmore2021]. Structural equation
modelling (SEM) and path analyses have been commonly used in ecology to
assess the causal effects behind various hypotheses [@fan2016;
@grace2010], including the relationship between forest structure,
functioning, and biodiversity [@ali2019]. For example, SEM has commonly
been used to assess the role of structural diversity of carbon storage
as compared to species diversity [@ali2016; @zhang2015].

The overall goal of this paper is to characterize patterns of forest
ecosystem structure and function and their interrelationships across a
wide range of ecosystems encompassing significant environmental
gradients. To do so, we synthesize data from moderate-scale
remote-sensing derived metrics of structure, represented as both simple
ALS metrics of canopy height and cover, as well as modeled forest
structure variables including volume, and above ground biomass, with a
remote sensing derived index on ecosystem function. Our first question
is to examine how ecosystem structure and function vary across a large
environmental gradient and then compare the simple and modeled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around redundancy in remote sensing metrics when describing
the structure and function of ecosystems and proposes a method to
examine potential overlap when generating remote sensing EBV's.

Our second question examines the independent and shared relationships of
ecosystem structure height and cover, with modeled forest structure, on
function. This provides insight into the choice of remote sensing
attributes to use when developing EBV's. Remote sensing datasets can
comprise relatively unprocessed, observations, in this case ALS measures
of height and cover which are retrieval from the raw 3D point cloud, vs
modeled attributes, such as biomass and volume, which involve the use
the statistical relationships to transform the observations into more
refined data products. Assessing which of these two (or combination of
the two) approaches has more or less redundancy with estimates of
function provides insights into the choice of data used to build EBV's.
Lastly we examine how the primary and modeled structure attributes vary
in their variance partitioning within key biomes across this large
environmental range, examining to what extend ecosystem type impacts
these interrelationships and thus the applicability of these results
globally.

@@

## Study Area

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
province is forested [@bcministryofforests2003]. There is a large amount
of ecosystem variation in the province, with large climate and
topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
system identifies 16 zones based on the dominant tree species and the
ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography [@pojar1987]. To examine trends across the large
environmental gradients, we group the BEC zones into five broad biomes,
specifically, the south interior, northern, montane, alpine, and coastal
groups similar to @hamann2006.

{insert figure showing ecosystems of interest} need to remake figure
showing ecosystems, highlighting that bunchgrass and alpine zones are
not included

{table showing elevations and climate ranges for each included bec zone}

@@

We utilize a suite of forest structural diversity variables (canopy
height, canopy cover, Lorey's height, overstory cover, basal area,
aboveground biomass, gross stem volume, mean elevation, elevation
standard deviation, and structural complexity \[coefficient of variation
in elevation returns\]). This dataset is created at a 30 m spatial
resolution according to @matasci2018. In brief, the method utilizes a
set of lidar collections and field plots across Canada, and imputes the
remaining pixels using a random forest k-Nearest Neighbour approach.
Detailed information on the creation of this dataset can be found in
@matasci2018.

@@

The Dynamic Habitat Indices are a set of satellite remote sensing
derived productivity variables that summarize the cumulative amount of
available energy, the minimum available energy, and the variation in
available energy throughout a given year [@radeloff2019]. The DHIs have
previously been produced at a global extent using MODIS imagery
[@radeloff2019], however, recent studies have examined how these indices
can be constructed at a finer spatial resolution using multi-annual
Landsat imagery.

The DHIs were calculated according to {RAZENKOVA LANDSAT DHI PAPER IN
PRESS} for all of terrestrial British Columbia. In brief, they utilized
Google Earth Engine [@gorelick2017] to obtain all valid Landsat pixels
for a given study area by filtering out low quality pixels, then
calculated the NDVI for each pixel. They use the median NDVI value for
each month across the ten year time span (2011-2020) to calculate the
sum, minimum, and coefficient of variation in NDVI values. More detailed
information can be found in {Razenkova In Press}.

@@

We conducted model based sampling across the fifteen forest dominated
ecosystems found within British Columbia \[BC figure reference, I have
not made this figure yet\]. Samples were randomly selected within BEC
zones of interest alongside multiple criteria. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in the Lorey's height and canopy cover forest structural
metrics. A maximum of 3000 samples were sampled in each BEC zone. All
variables were natural-log transformed and standardized. Variables
containing zeros were natural-log plus one transformed.

@@

### Partial Redundancy Analysis and Variation Partitioning

>>>>

---
title: Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
abstract: In today's changing climate and in light of the biodiversity crises, it is integral to be able to monitor the change in biodiversity at large scales. Key to this is the development of variables capable of monitoring biodiversity trends. One framework developed to tackle this problem is the Essential Biodviersity Variables (EBV), which is designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes. The EBV framework contains six classes, some of which are extremely well suited to be monitoring from satellite remote sensing. Among these with the potential to be satellite derived, the ecosystem structure and function classes stand out. In this paper, we seek to assess the complementarity of forest structure, as imputed across the entirety of British Columbia, Canada, with the Dynamic Habitat Indices, a yearly summary of productivity indices. Both datasets have been previously linked with biodiversity metrics across a range of scales. Using redundancy analysis, we find that forest structural attributes and the DHIs are decoupled from one another, with the forest structure datasets explaining 14% of the variation in the DHIs. Further, we explore how the proportion of variance explained in the DHIs from primary structural attributres such as canopy height and canopy cover and modelled structural attributes such as aboveground biomass varies. We find there is generally large overlap between the two forest structure datasets. Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV are well suited to be complementary EBVs across large environmental gradients.
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices
  - essential biodiversity variables
  - redundancy analysis
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Remote Sensing of Environment
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
---

@@

# include all libraries required by the R chunks here
library(vegan)
library(sgsR)
library(terra)
library(sf)
library(gt)
library(tidyverse)

@@

# generate package citations
loaded_pkgs = (.packages())
base_pkgs = getOption("defaultPackages")
contrib_pkgs = loaded_pkgs[!(loaded_pkgs %in% base_pkgs)]
pkgs_file = here::here("scripts", "manuscript", "packages.bib")
knitr::write_bib(contrib_pkgs, pkgs_file)

bec_tab <- read_csv(here::here("data", "climate.csv")) %>%
  mutate(across(.cols = c(tmax, tmin), \(x) round(x, digits = 1)))
```

@@

With biodiversity being in decline, and facing extinction rates above
the background extinction rate [@thomas2004; @urban2015], as well as the
homogenization of communities at various scales [@mcgill2015] it is
integral to be able to monitor how biodiversity is changing across the
globe. In response, the global biodiversity community is making efforts
to assess and halt the degradation of biodiversity. The Group for Earth
Observation Biodiversity Observation Network has developed the Essential
Biodiversity Variables [EBVs, @pereira2013], designed as an analog to
the Essential Climate Variables framework [@bojinski2014]. EBVs are
designed to be global in scope, relevant to biodiversity information,
feasible to use, and complementary to one another [@skidmore2021]. While
it can be incredibly difficult, time consuming, and expensive to collect
data on biodiversity across wide swaths of land and varying ecosystems,
EBVs, which can be correlated to sampled biodiversity information, allow
for the monitoring and assessment of protected area effectiveness and
ecosystem health at large spatial scales [@hansen2021]. There are six
EBV classes, each of which correspond to a different facet of
biodiversity including species populations, species traits, community
composition, ecosystem structure, ecosystem function, and genetic
composition [@pereira2013].

Satellite remote sensing has proven to be capable of measuring five of
the six EBV classes, the exception being genetic composition, which
requires in-situ observation and sampling [@skidmore2021]. Species
populations - and in turn community composition - can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale, however it is difficult and computationally expensive to extend
these analyses to broader extents [@fassnacht2016; @graves2016] while
species traits such as vegetation phenology have been observed at the
single-tree scale using, for example, PlanetScope imagery and
drone-based measurements [@wu2021]. However, the spatially limited and
often *ad-hoc* data collection approaches associated with monitoring
individuals is not conducive to the global or regional scales required
for biodiversity trend assessment [@valdez2023].

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer [MODIS; @zhang2003], the Landsat
imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@darvishzadeh2019] programs. These mid-resolution satellites can monitor
processes at broader extents but the coarse spatial resolution removes
the ability to relate these traits to individual organisms. As a result,
satellite remote sensing data has been shown to be arguably the most
effective at monitoring ecosystem based EBVs focused on structure and
function. These EBVs classes can be monitored at regional to global
extents through the use of optical imagery [@cohen2004], as well as
active sensors such as lidar (light detection and ranging) and radar
[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].

## EBV - Ecosystem Structure

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Structural attributes range in
complexity from simple (canopy cover; canopy height), to more complex
(vertical and horizontal structural complexity) to modelled (aboveground
biomass; basal area), all of which can be assessed using lidar data
[@coops2021]. A suite of these lidar-derived attributes have been used
as local indicators of biodiversity, including simple metrics such as
canopy cover and canopy height as well as derived metrics including
vertical profiles, aboveground biomass [@lefsky1999; @guo2017;
@coops2016]. Other second order derived metrics such as canopy texture,
height class distribution, edges, and patch metrics have also been used
to examine habitat and biodiversity at landscape scales [@bergen2009].
Advances in satellite remote sensing processing have allowed 3D forest
structure data to be imputed across wide spatial scales [@matasci2018;
@coops2021] using data fusion approaches involving collected lidar data
and optical/radar data.

Increased forest structural complexity has been hypothesized to create
additional niches, leading to increased species diversity [@bergen2009],
which has been frequently demonstrated using avian species diversity
metrics [@macarthur1961]. For example: @herniman2020 used spectral and
lidar derived forest structure data to model avian habitat suitability;
@clawges2008 found that lidar derived forest structural attributes are
capable of identifying habitat types associated with avian species in
pine/aspen forests; @goetz2007 used canopy structural diversity to
predict bird species richness, finding that canopy vertical distribution
was the strongest predictor of species richness. Forest structural
metrics have also been used to study biodiversity in other clades as
well [@davies2014; @nelson2005].

## EBV - Ecosystem Function

With respect to ecosystem function, energy availability in an ecosystem
has shown to be a predictor of species richness and abundances at
various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
and is measurable using satellite remote sensing via the use of various
vegetation indices [@huete2002; @radeloff2019]. Vegetation indices,
which are indicative of photosynthetic activity, are commonly used as
proxies of gross primary productivity [@huang2019]. These vegetation
indices have also been used to assess patterns in biodiversity at single
time points [@bonn2004], and more recently, through yearly summaries of
productivity [@berry2007; @radeloff2019]. The relationship between
energy availability and biodiversity occurs via various hypothesized
mechanisms, such as the available energy hypothesis [@currie2004;
@wright1983], the environmental stress hypothesis [@currie2004], and the
environmental stability hypothesis [@williams2008]. These three
hypotheses have in turn been linked to patterns of annual surface
reflectance in remote sensing data [@berry2007; @radeloff2019].

@berry2007 first explored this idea by proposing the linkage of
intra-annual summaries of MODIS derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices [DHIs;
@coops2008], which have now been shown to be well suited to assess the
three aforementioned hypotheses at global scales [@radeloff2019]. The
cumulative DHI calculates the total amount of energy available in a
given pixel over the course of a year. Cumulative DHI is strongly linked
to the available energy hypothesis, which suggests that with greater
available energy species richness will increase [@wright1983]. The
minimum DHI, which calculates the lowest productivity over the course of
a year can be matched to the environmental stress hypothesis, which
proposes that higher levels of minimum available energy will lead to
higher species richness [@currie2004]. Finally, the variation DHI, which
calculates the coefficient of variance in a vegetation index through the
course of a year, corresponds to the environmental stability hypothesis
which states that lower energy variation throughout a year will lead to
increased species richness [@williams2008].

## Biodiversity Monitoring with EBVs

Biodiversity monitoring programs often require a range of information in
order to accurately assess changes in ecosystem integrity
[@lindenmayer2010]. Choosing datasets that are most closely related to
the phenomenon of interest in a given application allows for direct
connections to monitoring results and management actions [@pressey2021].
With the advent of large-extent monitoring methods like satellite remote
sensing, and a proliferation of potential EBVs datasets, it becomes
important to assess the interrelationships between these datasets, and
assess their complementary of the information to reduce duplication of
efforts [@pereira2013; @skidmore2021]. When strong relationships are
present between EBV classes, it becomes possible to assess the
ecological relationships between potential EBVs. On the other hand, when
datasets do not appear related, they may be well suited to be used in
monitoring programs together, as complementary EBVs.

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for over 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(within a single image) [@myneni1994], various relationship directions
and shapes have been found between forest structure and function metrics
[@ali2019]. Hypothesized mechanisms such as niche complementary have
shown that aboveground biomass increases with stand structure
[@zhang2012], while asymmetric competition for light can reduce forest
productivity with increased structural complexity [@bourdier2016]. The
relationship in particular between forest structural diversity metrics -
which are now more accurately and comprehensively derived from lidar
data - and temporal variation in functional metrics, specifically the
metrics of ecosystem productivity via the DHI framework, have yet to be
fully examined.

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function and their complementarity across a wide range of
ecosystems encompassing significant environmental gradients. To do so,
we synthesize data from moderate-scale remote-sensing derived metrics of
forest structure, represented as both simple ALS-extracted metrics of
canopy height, cover and vertical complexity, as well as modelled forest
structure attributes including volume, and aboveground biomass, with a
well-established remote sensing derived index on ecosystem function. Our
first question is to examine how ecosystem structure and function
complement one another across a large environmental gradient and then
compare the simple and modelled representations of forest structure to
different levels of ecosystem function. This question is important as it
provides insights to the EBV community around complementarity of remote
sensing metrics when describing the structure and function of ecosystems
and proposes a method to examine potential overlap when generating
remote sensing EBVs.

Our second question examines the independent and shared relationships of
ecosystem structure height and cover, with modelled forest structure, on
ecosystem function. This provides insight into the choice of remote
sensing attributes to use when developing EBVs within a single EBV
class. Remote sensing datasets can comprise relatively unprocessed
observations, in this case ALS measures of height and cover which are
derived from the raw 3D point cloud vs modelled attributes, such as
biomass and volume, which involve the use the statistical relationships
with field data to transform the observations into more refined data
products. Assessing which of these two (or combination of the two)
approaches has stronger or weaker correlations with estimates of
function provides insights into the choice of data used to build EBVs.
Lastly, we examine how the primary and modelled structure attributes
partition the variance of the DHIs within key biomes and forest types
across a large environmental range, examining to what extent ecosystem
and forest types impacts these relationships and thus providing insight
into the applicability of these results globally.

@@

## Study Area {#sec-study-area}

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems. Approximately 64% of the province is
forested, with large environmental and topographic gradients
[@pojar1987; @bcministryofforests2003]. The Biogeoclimatic Ecosystem
Classification (BEC) system identifies 16 zones based on the dominant
tree species and the ecosystems general climate (@fig-study-map).

```{r}
#| label: fig-study-map
#| fig-cap: Location of biogeoclimatic ecosystem classification (BEC) zones in British Columbia.

knitr::include_graphics(here::here("outputs", "bec_map.png"))
```

These zones can be further split into subzones, variants, and phases
based on microclimate, precipitation, and topography [@pojar1987]. To
examine trends across the large environmental gradients, we group the
BEC zones into five broad biomes, specifically, the southern interior,
northern interior, montane, alpine, and coastal groups similar to
@hamann2006. We also report each BEC zone's average climate data,
according to @wang2020 (@tbl-bec-group).

```{r}
#| label: tbl-bec-group
#| tbl-cap: BEC Zones, their aggregate groups, and their average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016.

# bec_tab %>%
#   gt() %>%
#   cols_label(
#     group = "BEC Group",
#     zone = "BEC Zone",
#     zone_nm = "Full Name",
#     ppt = "Precipitation (mm/yr)",
#     tmax = "Max Temperature (°C)",
#     tmin = "Min Temperature (°C)"
#   ) %>%
#   as_word()

knitr::kable(
  bec_tab,
  col.names = c(
    "BEC Group",
    "BEC Zone",
    "Full Name",
    "Precipitation (mm/yr)",
    "Max Temperature (°C)",
    "Min Temperature (°C)"
  )
)
```

@@

We used a suite of forest structural attributes (canopy height, canopy
cover, overstory cover, basal area, aboveground biomass, gross stem
volume, mean elevation, elevation standard deviation, and structural
complexity \[coefficient of variation in elevation returns\]). This
dataset was created for the year 2015 at a 30 m spatial resolution
according to @matasci2018. In brief, the method used a set of lidar
collections and field plots across Canada, and imputed the remaining
pixels using a random forest k-Nearest Neighbour approach on
Landsat-derived surface reflectance and auxiliary data such as
topography. Detailed information on the creation of this dataset can be
found in @matasci2018.

@@

We use an established set of indices of annual productivity shown to be
related to global biodvierstiy trends: the Dynamic Habitat Indices
[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
productivity variables that summarize the cumulative amount of available
energy, the minimum available energy, and the variation in available
energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
have previously been produced at a global extent using MODIS imagery ,
and have been used to assess alpha [@radeloff2019] and beta[@andrew2012]
diversity, species abundances [@razenkova2023], and construct novel
ecoregionalizations [@coops2009]; @andrew2013\] . Recent studies have
began to examine how these indices can be constructed at a finer spatial
resolution by using multi-annual Landsat imagery to generate a single
synthetic year of monthly observations [@razenkova2022].

The DHIs were calculated according to [@razenkova2022; Razenkova et al.,
In Press] for all of terrestrial British Columbia. In brief, Google
Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
for a given study area, filtering out pixels containing shadows, slouds,
and cloud shadows within each image [@zhu2012], then calculated the NDVI
for each pixel in each image. They then calculated the median NDVI value
for each month across the ten year time span (2011-2020) to generate a
synthetic year of monthly data. The sum, minimum, and coefficient of
variation across this synthetic year of NDVI values is then calculated.
More detailed information can be found in Razenkova et al. (In Press).

@@

We conducted model-based sampling across the sixteen ecosystems found
within British Columbia (@fig-study-map). Samples were randomly selected
within each BEC zone, in undisturbed pixels. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class. The land
cover mask was generated following [@hermosilla2022] using a
best-available-pixel composite, and an inverse-distance weighted random
forest approach across Canada. Additionally, each pixel had to have a
coefficient of variation less than 0.5 in surrounding pixels in the two
simplest forest structural attributes, canopy height and canopy cover. A
maximum of 3000 samples were sampled in each BEC zone with a 1 km
minimum sampling distance to reduce the effects of spatial
autocorrelation. All variables were natural-log transformed and
standardized. Variables containing zeros were natural-log plus one
transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
classes and coefficient of variation were calculated in Python version
3.9.

# Sampling 2

We implement two sampling schemes -- one for ecosystems, and one for
forest types -- across the terrestrial ecosystems of British Columbia
(@fig-study-map). Both have samples randomly selected. For both sampling
schemes, each individual sample was taken from a forested pixel which
was surrounded by the same land cover class. The land cover class map
was generated for the year 2015 following [@hermosilla2022] using a
best-available-pixel composite, and an inverse-distance weighted random
forest approach across Canada. Additionally, each pixel had to have a
coefficient of variation less than 0.5 in surrounding pixels in the two
simplest forest structural attributes, canopy height and canopy cover. A
minimum sampling distance of 1 km was implemented to reduce the effects
of spatial autocorrelation, and samples that had been disturbed in the
last 30 years were discarded by using a disturbance mask generated for
the forested ecosystems of Canada by @hermosilla2016.

The ecosystem sampling scheme was implemented by randomly sampling
pixels matching the above criteria within each BEC zone of British
Columbia, up to a maximum of 3000 pixels, regardless of land cover
class. The forest type sampling scheme was similar in that each BEC zone
was sampled, however, it implemented stratified sampling on each forest
type, up to a maximum of 500 pixels per BEC zone and forest type.

Each sampling scheme was natural-log transformed and standardized to
Z-scores. Variables containing zeros were natural-log plus one
transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
classes and coefficient of variations of canopy height and cover were
calculated in Python version 3.9.

@@

### Redundancy Analysis and Variation Partitioning

Redundancy analysis (RDA) and variation partitioning were used to relate
the primary and modelled forest structure attributes to ecosystem
function across a broad environment range. Redundancy analysis functions
similarly to a multiple linear regression, except it is capable of
predicting multiple response variables. It accomplishes this by first
running a multiple linear regression of each predictor variable on each
response variable, then running a principle component analysis on the
residuals from each multiple linear regression. This reduces the
dimensionality of the output, and allows the relationship strength to be
assessed by calculating the loadings of both predictor and response
variables on the RDA axes. Partial redundancy analysis function
similarly, except also considers co-variates [@legendre2012]. Redundancy
analysis has widely been used in community ecology where environmental
variables of interest are compared to species composition
[@blanchet2014; @kleyer2012] and has similarities to partial least
squares regression -- namely, their multivariate approach, usage of
dimensionality reduction, and linearity assumptions -- which is commonly
used in remote sensing literature [@roelofsen2014; @burnett2021].

Following the RDA, we employ ANOVAs to determine which axes are
significant, and calculate the proportion of variance attributable to
each axis using the eigenvalues generated from the RDA. We calculate
axis loadings for both predictor and response variables by calculating
the correlation between the variables and the RDA axes. Axis loadings
represent the relationship between a given variable and the RDA axis. We
only consider and display significant axes. To visualize the RDA for
both predictor and response variables, we display the results as path
diagrams, with loadings from each predictor to the RDA axis to the
response variables. The variance explained by each axis is also
displayed in the RDA box (@fig-rda-var). All RDA calculations were done
in R [@R-base] version 4.2.2 using the **vegan** package [@R-vegan].
